\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-05-09 docs/papers/sto/so.tex
Updating paper draft

\begin{figure*}
  \centering
  \includegraphics[width=.98\columnwidth]{slab_s0_snap120_3177874_e3000.pdf}
  \caption{Projected dark matter density fields of \GADGET, original PM,
  and optimized PM-SONN.
  The examples here are shown for the 0th data point in our samples of
  configurations and cosmological parameters.
  This corresponds to an overall simulation box size of 79 Mpc$/h$.
  We show a zoom-in view with window length XXX Mpc$/h$, and a projected
  slab thickness of XXX Mpc$/h$.}
  \label{fig:denslab}
\end{figure*}

@@

@@

\begin{table}
  \centering
  \caption{Possible features that can affect force sharpening.
  All scales are then multiplied by the wavenumber $k$ to form dimensionless
  factors.}
  \label{tab:feat}
  \begin{tabular}{rl}
  \toprule
  type & feature \\
  \midrule
  & $R_\mathrm{d}$: RMS linear displacement \\
  \cmidrule(lr){2-2}
  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $\d R / \d\lna$: time derivatives of the above \\
  \midrule
  & particle spacing \\
  \cmidrule(lr){2-2}
  other scales & $l$: PM force mesh cell size \\
  \cmidrule(lr){2-2}
  & softening length \\
  \midrule
  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln G_m / \d\lna$ \\
  \cmidrule(lr){2-2}
  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln\!H / \d\lna$ \\
  \cmidrule(lr){2-2}
  & $\Delta\lna$: time step size in $\lna$ \\
  & \YL{Now this can be an important feature with varying $a_\ic$} \\
  \bottomrule
  \end{tabular}
  \end{table}

We design the input features of the NNs in a way such that they could
properly fit and interpolate given our wide coverage of the
configurations and cosmological parameters.

Multiple scales and dimensionless factors could possibly affect force
sharpening, which are denoted as $\vvartheta$ and enter the NNs of $f(k_i;
\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ as input features.
A list of these features is summarized in \autoref{tab:feat}, and here we
present relevant definations and more details.

Below are some definations related to the typical scales.
\begin{itemize}
\item The root-mean-square (RMS) linear displacement is given by
%
\begin{equation}
R_\mathrm{d} = \int \frac{k P_\linear(k)}{6\pi^2} \d\lnk,
\end{equation}
%
where $P_\linear(k)$ is the linear power spectrum.
\item The dimensionless linear power spectrum is defined as
%
\begin{equation}
\Delta^2_\linear(k) \triangleq \frac{k^3 P_\linear(k)}{2 \pi^2},
\end{equation}
%
based on which we define a nonlinear scale $R_P$ through
$\Delta_\linear^2(R_P^{-1}) \triangleq 1$.
\item The variance of linear matter overdensity in a top-hat window is given by
%
\begin{equation}
\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
  \Delta_\linear^2(k) W_\tophat^2(kR),
\end{equation}
%
where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
Fourier transform of the top-hat window.
Then we have $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$.
\item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$: $\sigma_\gauss^2(R_\gauss)
\triangleq 1$.
\end{itemize}
These nonlinear scales are functions of time since the linear power
spectrum evolves with time through the growth factor $D(a)$, we also
take the time derivatives of these scales as input features.
Other relevant scales include the particle spacing in Lagrangian space,
the cell size of the chosen mesh grid for force evaluation, and the
softening length in the target \GADGET\ simulations.

@@

Dimensionless combinations are formed by multiplying the wavevector
components with the scales, i.e.,
%
\begin{equation}
f(k_i; \vvartheta) = f(k_iR_d, k_iR_P, \cdots; G_m, \Omegam, \cdots),
\end{equation}
%
\begin{equation}
g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_P, \cdots\}_{j=1,2,3};
                                 G_m, \Omegam, \cdots).
\end{equation}
%

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=.98\columnwidth]{slab_s0_snap120_3177874_e3000.pdf}
  \caption{Projected dark matter density fields of \GADGET\
  (\textit{middle}), original PM (\textit{left}), and optimized PM-SONN
  (\textit{right}).
  The examples here are shown for the 0th data point in our samples of
  configurations and cosmological parameters.
  This corresponds to an overall simulation box size of 79 Mpc$/h$.
  We show a zoom-in view with window length XXX Mpc$/h$, and a projected
  slab thickness of XXX Mpc$/h$.}
  \label{fig:denslab}
\end{figure*}

@@

We design the input features of the NNs and SR in a way such that they
could properly fit and interpolate given our wide coverage of the
configurations and cosmological parameters.
Given the difference between NNs and SR, we could have different input
features for the two.
When generating the data for SR with the trained NNs, we could reduce or
add more features based on the existing features, with the output
unchanged.

@@

\begin{table}
  \centering
  \caption{Possible features that can affect force sharpening, detailed
  in \autoref{sec:features}. Scales could be multiplied by the
  wavenumber $k$ to form dimensionless factors.}
  \label{tab:feat}
  \begin{tabular}{rl}
  \toprule
  type & feature \\
  \midrule
  & $R_\mathrm{d}$: RMS linear displacement \\
  \cmidrule(lr){2-2}
  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
  \cmidrule(lr){2-2}
  & $\d R / \d\lna$: time derivatives of the above \\
  \midrule
  & $l_p$: particle spacing \\
  \cmidrule(lr){2-2}
  other scales & $l_c$: PM force mesh cell size \\
  \cmidrule(lr){2-2}
  & $l_s$: softening length \\
  \midrule
  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln G_m / \d\lna$ \\
  \cmidrule(lr){2-2}
  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
  \cmidrule(lr){2-2}
  & $\d\ln\!H / \d\lna$ \\
  \cmidrule(lr){2-2}
  & $\Delta\lna$: time step size in $\lna$ \\
  & \YL{Now this can be an important feature with varying $a_\ic$} \\
  \bottomrule
  \end{tabular}
  \end{table}

In this section, we introduce a list of physical quantities that could
be of possible importance in SO.
These quantities are used in the input features for either the NNs or
SR.
Multiple scales and dimensionless factors could possibly affect force
sharpening.
A list of these features is summarized in \autoref{tab:feat}, and here
we present relevant definitions and more details.

Below are some definitions related to the typical scales.
\begin{itemize}
\item The root-mean-square (RMS) linear displacement is given by
%
\begin{equation}
R_\mathrm{d} = \int \frac{k P_\linear(k)}{6\pi^2} \d\lnk,
\end{equation}
%
where $P_\linear(k)$ is the linear power spectrum.
\item The dimensionless linear power spectrum is defined as
%
\begin{equation}
\Delta^2_\linear(k) \triangleq \frac{k^3 P_\linear(k)}{2 \pi^2},
\end{equation}
%
based on which we define a nonlinear scale $R_P$ through
$\Delta_\linear^2(R_P^{-1}) \triangleq 1$.
\item The variance of linear matter overdensity in a top-hat window is given by
%
\begin{equation}
\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
  \Delta_\linear^2(k) W_\tophat^2(kR),
\end{equation}
%
where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
Fourier transform of the top-hat window.
Then we have $R_\tophat$ defined through $\sigma_\tophat^2(R_\tophat)
\triangleq 1$.
\item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$ with
$\sigma_\gauss^2(R_\gauss) \triangleq 1$.
\end{itemize}
These nonlinear scales are functions of time since the linear power
spectrum evolves with time through the growth factor $D(a)$, we could
also take the time derivatives of these scales as possible features.
Other relevant scales include the particle spacing $l_p$ in Lagrangian
space, the cell size $l_c$ of the chosen mesh grid for force evaluation,
and the softening length $l_s$ in the target \GADGET\ simulations.

@@

Scales could be multiplied with the wavenumber $k$ to form dimensionless
quantities.
For example, the product of $k$ with the mesh cell size $l_c$ results in
a quantity $kl_c = \pi k/k_{\rm Nyquist}$ that is invariant to different
box sizes, which should be an important feature.
